BOLD Parameter Estimation Using Particle Filters

### 2010-07-21 doc/thesis/background.tex
removed lots of weasel words...

Currently, FMRI is used to determine the location of responses
to stimuli. The method of finding activation is discussed in 
\autoref{sec:Statistical Parametric Mapping}. The goal of this 
work is to move away from the question "Is this region active"
and instead ask "What does the activation look like". Answering
this question necessitates more complex models and certainly
will result in longer run times. Already there have been several other
attempts to model the BOLD response; these works will be discussed
in in the next few sections.

@@

As mentioned previously, a square wave stimulus 
does not result in a square wave in the activation of brain regions. 
The BOLD signal is in fact a significantly smoothed version of the 
stimuli. As such, when fitting an FMRI time course to the input,
the input ($X$'s columns) is usually smoothed to reduce bias error. 
The best method, that maintains a linear fit, is convolving the input 
with a Hemodynamic 
Response Function (HRF). The \emph{Hemodynamic Response Function}
mimics the basic shape of BOLD activation, including a delay
due to rise time and fall time. The fitting 
process is then a least squares fit over the space of the vector $\beta$. 
Therefore $Y(t)$ is estimated as a linear combination 
of the columns of $X$.

@@

\subsection{Discussion}
\label{sec:BackgroundConclusion}
In all, the GLM is extremely useful for determining linear 
dependence of a set of regressors on the output. Unfortunately, as discussed in
\autoref{sec:BOLD Analysis} there are significant nonlinearities 
that almost certainly cause false negatives in the Statistical Parametric
Maps. Unfortunately nonlinear analyses have only recently become feasible,
so the scale of the problem is still unknown. The problem is 
highlighted by the relatively common scenario where no significant
activation can be found in a single run \cite{Riera2004}
\cite{Johnston2008}.

@@

\section{Solving the BOLD Model}
Unlike Statistical Parametric Mapping, the techniques described in this
section are all attempts to learn some version of the BOLD model. 
There have been quite a few efforts to quantify the parameters of the
various BOLD models.  Although \cite{Buxton1998} and \cite{Friston2000}
both proposed physiologically reasonable values for the model parameters, 
\cite{Friston2002b} was the first paper to calculate the parameters based 
on actual FMRI data. However, in that case, the voxels were chosen from
regions that were detected as active by the GLM. It is therefore possible
that the parameters are biased toward parameters that fit the linear model.

\subsection{Linear Approximation}
In \cite{Friston2002b}, a novel combination of linear and nonlinear modeling
was used to generate parameter estimates. Because it is impossible to 
calculate the partial derivative of
the output with respect to parameters, $\frac{\partial \theta}{\theta}$, 
\cite{Friston2002b} approximates the partial. At each step of the Expectation-
Maximization algorithm, the differential equation is integrated, calculating
the residuals. Then, for each $\theta_i$ surrounding the current estimate of 
$\theta$, a Volterra-Kernel expansion of the output $y$ is generated. Generation
of the Volterra Kernel is relatively quick, and, since it is linear there
is an analytical solution. Thus, once the Volterra Kernel exists, the value
of $y$ at that measurement point, for that $\theta_i$ can easily be found. 
Thus, one part of the partial is calculated at one point using
$\frac{y(t) - y_i(t)}{\theta - \theta_i}$. This is repeated for
every dimension of $\theta$ to give the full $\frac{\partial y}{\theta}(t)$.
This is only at time $t$ though, and it must be repeated for every
measurement. Finally the full derivate matrix is filled out, and
the next step in the E-M algorithm can proceed. The full E-M algorithm
for estimating the states is notation-heavy and can be found in \cite{Friston2002b}.

@@

Although there are certainly benefits to using a derived model, rather
than a purely empirical model, there are serious implications. The
first problem is that all the powerful classical techniques of gradient
descent are off limits; since the model is a true nonlinear dynamical
system with no closed form solution. The implication of this is that
the calculation of a Jacobian for residuals won't work; and thus powerful
techniques such as the Gauss-Newton method, which are helpful in many nonlinear
problems, are off limits. Additionally, a gradient descent is difficult to
perform without the ability to form partials of the output with respect
to all the parameters.

@@

Another potential method of interest is the use of Genetic Algorithms
(GA). Genetic algorithms are similar to Simulated Annealing, in
that they randomly move to better solutions based on a cost function.
However; in genetic algorithms a single point estimate isn't used. Instead
a population of estimates is generated, each with distinct parameters,
and then each set of parameters is rated with a fitness function. Parameter
sets that are good get a higher weight; then new parameter sets are generated by 
randomly combining pieces of the old parameter sets. The pieces are typically
chosen at a rate proportional to the fitness of the donor; thus fit
parameter sets tend to pass on their properties. In addition to this,
random mutations may be introduced that come from no existing parent. 
The new generation is then rated with the fitness function again, and
the entire process starts over. The stop condition for a genetic algorithm
is typically based on some threshold for fitness or a maximum number 
of generations. This is problematic, since it doesn't really provide 
any guarantee of even reaching a local minima, although in practice
it can be quite effective.

%[genetic algorithm picture] todo

@@

In order to estimate the posterior at $t$, a deterministic set of sigma points 
(often 2 per dimension, plus 1 at the mode of the multivariate distribution)
weighted according to a Gaussian estimate of $X(t-1)$ are passed through
the update equation. This set of points are then used to estimate the 
mean and covariance of $X(t)$. The benefit of this, is that it requires
no Jacobian and only a few extra calculations to get a decent esimate of
a posterior Gaussian. In the BOLD case, the set of equations we are modeling have no 
closed form solution, and finding the Jacobian is impossible without approximations. Although 
\cite{Riera2004}, \cite{Hu2009} mention a Jacobian of the BOLD response, this is
not strictly the case and is rather $\frac{\partial J}{\partial t}$
rather than a true Jacobian.  This is important because the Extended
Kalman filter depends on the Jacobian to map a Gaussian through the 
advancement of time. Thus the popular Extended Kalman Filter won't work
in this case, whereas the Unscented Kalman Filter still does. In fact
\cite{Hu2009} uses the UKF to perform a similar type of analysis to
the one performed in this work.

@@

The difficulty
of using a Kalman Filter, however, is that it assumes a multivariate 
Gaussian for the state variables, $X(t-1)$. The problem with this is that 
when a system is significantly nonlinear, the state at 
$X(t)$ will almost certainly be non-Gaussian, and thus estimating
$X(t+1)$ with $X(t)$ as a multivariate Gaussian will perpetually introduce
error in the distribution. Furthermore, it is not really known what 
sort of underlying distributions may exist in such a mixed biological,
mechanical, chemical system such as the brain. Assuming the parameters
all to be Gaussian may in fact be a gross error. On the other hand, for
small variances and short time steps the gaussiann distribution is a good 
fit, and so in some limited cases the Unscented Kalman Filter could work
well. These are 
non-trivial issues given that the assumption of Gaussianity is what allows
the UKF to estimate the posterior using only the first and second moments;
two parameters that don't uniquely describe most distributions.

@@

Notably $s_t$ has intentionally been set to a non-equilibrium, but physiologically
plausible value. The value of $u$ is left at zero the entire time, so the 
system will decay naturally (see \autoref{sec:BOLD Physiology}), though initializing 
$s$ at a non-zero level will drive the system for several seconds. 
\autoref{fig:transp1s} shows the results when the system is essentially
left on for 100 milliseconds after setting the variables according to \autoref{tab:steptable}.
The Q-Q plots fit very will with a Gaussian, demonstrating that at this short
time interval nonlinearities have not yet begun to effect the distribution.
However, \autoref{fig:trans1s} us the result after 1 second, which is faster
than most FMRI scanners are capable of sampling at. At that range the tails 
of the distributions for $v$ and $q$ are clearly starting to deviate from the
Gaussian distribution. As a result the uncertainty in $y$ is deviating from
the Gaussian distribution as well. This is important, because although 
approximating the distribution with a Gaussian based on the first two
moments will work in the short run, there will be residual error in the
distribution.

@@

Another potential problem with the UKF is that typically the sigma points,
used to estimate the posterior probability, $P(X(t) | X(t-1), u(t))$, are
located on the main axes. As a result, the covariances are not allowed to 
be effected by the state transition the same way the variances are. While
this may be reasonable in low-dimensional systems, high dimensional systems
have a much greater potential for interplay between the variables. While
this problem is relatively easy to fix (by using more sigma points off
the main axes), there is a definite cost in complexity.

@@

In \cite{Johnston2008}, a hybrid particle filter/gradient
descent algorithm was used to simultaneously derive the static and dynamic 
parameters, (classically known as parameters and state variables, respectively).
Essentially a particle filter is used to calculate the state variables at each
time; then the estimated distribution of the particles is used to find
the most likely set of parameters that would give that distribution of state variables.
This process is repeated until the parameters converge. Interestingly
\cite{Johnston2008} comes to a very different set of parameter estimates as compared
to the original \cite{Friston2000} estimates (\autoref{tab:Params}). In fact the results 
are significantly different from virtually every other study. The most obvious discrepancy
is the significantly longer time constants, $\tau_f$, $\tau_s$ and $\tau_0$. 
While of course this could be poor convergence of the algorithm, there is another other possibility.
Unlike all the other methods mentioned, excepting the methods in \autoref{sec:Nonlinear Least Squares},
the algorithm described in \cite{Johnston2008} does not depend on prior distributions.
It is possible then that the bias toward the prior in other methods screwed the results. 
While \cite{Johnston2008} is certainly in the minority; further exhaustive studies 
of the parameters, using unbiased techniques may be called for. A further 
comparison between the distributions found in \cite{Johnston2008} and 
\cite{Friston2000} will be discussed in \autoref{sec:PriorDist}, including
graphs: \autoref{fig:MeanResponseF}, \autoref{fig:MeanResponseJ}.

\section{Conclusion}
Clearly there is no ideal solution to solving this system of nonlinear
equations. Exhaustive search type methods such as those employed by
\cite{Johnston2008} and \cite{Vakorin2007} can have extremely long
run times even for a single voxel. While Volterra models are an interesting
solution,there have not yet been exhaustive tests to determine whether
such approximations work well throughout state space. The most promising
method of those reviewed here is the Kalman filter based method. It is
able to maintain a very fast runtime while still approaching the solution.
While the reliance on a Gaussian estimate to the true posterior distribution could
cause problems, some modifications could make it extremely powerful. 
The particle filter method proposed in the next section bears a strong 
resemblance to the unscented Kalman filter; albeit with more point estimating
the posterior.

>>>>

Currently, FMRI is used to determine the location of responses
to stimuli. The method of finding activation is discussed in 
\autoref{sec:Statistical Parametric Mapping}. The goal of this 
work is to move away from the question "Is this region active"
and instead ask "What does the activation look like". Answering
this question necessitates more complex models and certainly
will result in longer run times. Already there have been several other
attempts to model the BOLD response; these works will be discussed
in this chapter.

@@

As mentioned previously, a square wave stimulus 
does not result in a square wave in the activation of brain regions. 
The BOLD signal is in fact a smoothed version of the 
stimuli. As such, when fitting an FMRI time course to the input,
the input ($X$'s columns) is usually smoothed to reduce bias error. 
The best method, that maintains a linear fit, is convolving the input 
with a Hemodynamic 
Response Function (HRF). The \emph{Hemodynamic Response Function}
mimics the basic shape of BOLD activation, including a delay
due to rise time and fall time. The fitting 
process is then a least squares fit over the space of the vector $\beta$. 
Therefore $Y(t)$ is estimated as a linear combination 
of the columns of $X$.

@@

\subsection{Discussion}
\label{sec:BackgroundConclusion}
In all, the GLM is useful for determining linear 
dependence of a set of regressors on the output. Unfortunately, as discussed in
\autoref{sec:BOLD Analysis} there are significant nonlinearities 
that almost certainly cause false negatives in the Statistical Parametric
Maps. Unfortunately nonlinear analyses have only recently become feasible,
so the scale of the problem is still unknown. The problem is 
highlighted by the common scenario where no significant
activation can be found in a single run \cite{Riera2004}
\cite{Johnston2008}.

@@

\section{Solving the BOLD Model}
Unlike Statistical Parametric Mapping, the techniques described in this
section are all attempts to learn some version of the BOLD model. 
Although \cite{Buxton1998} and \cite{Friston2000}
both proposed physiologically reasonable values for the model parameters, 
\cite{Friston2002b} was the first paper to calculate the parameters based 
on actual FMRI data. However, in that case, the voxels were chosen from
regions that were detected as active by the GLM. It is therefore possible
that the parameters are biased toward parameters that fit the linear model.

\subsection{Linear Approximation}
In \cite{Friston2002b}, a novel combination of linear and nonlinear modeling
was used to generate parameter estimates. Because it is impossible to 
calculate the partial derivative of
the output with respect to parameters, $\frac{\partial \theta}{\theta}$, 
\cite{Friston2002b} approximates the partial. At each step of the Expectation-
Maximization algorithm, the differential equation is integrated, calculating
the residuals. Then, for each $\theta_i$ surrounding the current estimate of 
$\theta$, a Volterra-Kernel expansion of the output $y$ is generated. Generation
of the Volterra Kernel is quick, and, since it is linear there
is an analytical solution. Thus, once the Volterra Kernel exists, the value
of $y$ at that measurement point, for that $\theta_i$ can easily be found. 
Thus, one part of the partial is calculated at one point using
$\frac{y(t) - y_i(t)}{\theta - \theta_i}$. This is repeated for
every dimension of $\theta$ to give the full $\frac{\partial y}{\theta}(t)$.
This is only at time $t$ though, and it must be repeated for every
measurement. Finally the full derivate matrix is filled out, and
the next step in the E-M algorithm can proceed. The full E-M algorithm
for estimating the states is notation-heavy and can be found in \cite{Friston2002b}.

@@

Although there are certainly benefits to using a derived model, rather
than a purely empirical model, there are serious implications. The
first problem is that all the powerful classical techniques of gradient
descent are off limits; since the model is a true nonlinear dynamical
system with no closed form solution. The implication of this is that
the calculation of a Jacobian for residuals won't work; and thus powerful
techniques such as the Gauss-Newton method, which are helpful in most nonlinear
problems, are off limits. Additionally, a gradient descent is difficult to
perform without the ability to form partials of the output with respect
to all the parameters.

@@

Another potential method of interest is the use of Genetic Algorithms
(GA). Genetic algorithms are similar to Simulated Annealing, in
that they randomly move to better solutions based on a cost function.
However; in genetic algorithms a single point estimate isn't used. Instead
a population of estimates is generated, each with distinct parameters,
and then each set of parameters is rated with a fitness function. Parameter
sets that are good get a higher weight; then new parameter sets are generated by 
randomly combining pieces of the old parameter sets. The pieces are typically
chosen at a rate proportional to the fitness of the donor; thus fit
parameter sets tend to pass on their properties. In addition to this,
random mutations may be introduced that come from no existing parent. 
The new generation is then rated with the fitness function again, and
the entire process starts over. The stop condition for a genetic algorithm
is typically based on some threshold for fitness or a maximum number 
of generations. As with all nonlinear searches, unless the function is
known to be convex, there is no guarantee that a global minimum has been
reached.
%[genetic algorithm picture] todo

@@

In order to estimate the posterior at $t$, a deterministic set of sigma points 
(often 2 per dimension, plus 1 at the mode of the multivariate distribution)
weighted according to a Gaussian estimate of $X(t-1)$ are passed through
the update equation. This set of points are then used to estimate the 
mean and covariance of $X(t)$. The benefit of this, is that it requires
no Jacobian and only a few extra calculations to get a decent estimate of
a posterior Gaussian. In the BOLD case, the set of equations we are modeling have no 
closed form solution, and finding the Jacobian is impossible without approximations. Although 
\cite{Riera2004}, \cite{Hu2009} mention a Jacobian of the BOLD response, this is
not strictly the case and is rather $\frac{\partial J}{\partial t}$
rather than a true Jacobian.  This is important because the Extended
Kalman filter depends on the Jacobian to map a Gaussian through the 
advancement of time. Thus the popular Extended Kalman Filter won't work
in this case, whereas the Unscented Kalman Filter still does. In fact
\cite{Hu2009} uses the UKF to perform a similar type of analysis to
the one performed in this work.

@@

The difficulty
of using a Kalman Filter, however, is that it assumes a multivariate 
Gaussian for the state variables, $X(t-1)$. The more nonlinear the system
gets the more likely that the Gaussian will be insufficient to describe
the distribution, $X(t)$. When this occurs, every step from  $X(t+1)$ 
to $X(t)$ will introduce additional
error in the posterior distribution. Furthermore, it is not really known what 
sort of underlying distributions may exist in such a mixed biological,
mechanical, chemical system such as the brain. Assuming the parameters
all to be Gaussian may in fact be a gross error. On the other hand, for
small variances and short time steps the Gaussian distribution is a good 
fit, and so in some limited cases the Unscented Kalman Filter could work
well. These are 
non-trivial issues given that the assumption of Gaussianity is what allows
the UKF to estimate the posterior using only the first and second moments;
two parameters that don't uniquely describe most distributions.

@@

Notably $s_t$ has intentionally been set to a non-equilibrium, but physiologically
plausible value. The value of $u$ is left at zero the entire time, so the 
system will decay naturally (see \autoref{sec:BOLD Physiology}), though initializing 
$s$ at a non-zero level will drive the system for several seconds. 
\autoref{fig:transp1s} shows the results when the system is 
left on for 100 milliseconds after setting the variables according to \autoref{tab:steptable}.
The Q-Q plots fit will with a Gaussian, demonstrating that at this short
time interval nonlinearities have not yet begun to effect the distribution.
However, \autoref{fig:trans1s} us the result after 1 second, which is faster
than most FMRI scanners are capable of sampling at. At that range the tails 
of the distributions for $v$ and $q$ are starting to deviate from the
Gaussian distribution. As a result the uncertainty in $y$ is deviating from
the Gaussian distribution as well. This is important, because although 
approximating the distribution with a Gaussian based on the first two
moments will work in the short run, there will be residual error in the
distribution.

@@

Another potential problem with the UKF is that typically the sigma points,
used to estimate the posterior probability, $P(X(t) | X(t-1), u(t))$, are
located on the main axes. As a result, the covariances are not allowed to 
be effected by the state transition the same way the variances are. While
this may be reasonable in low-dimensional systems, high dimensional systems
have a much greater potential for interplay between the variables. While
this problem is simple to fix (by using more sigma points off
the main axes), there is a definite cost in complexity.

@@

In \cite{Johnston2008}, a hybrid particle filter/gradient
descent algorithm was used to simultaneously derive the static and dynamic 
parameters, (classically known as parameters and state variables, respectively).
A particle filter is used to calculate the state variables at each
time; then the estimated distribution of the particles is used to find
the most likely set of parameters that would give that distribution of state variables.
This process is repeated until the parameters converge. Interestingly
\cite{Johnston2008} comes to a very different set of parameter estimates as compared
to the original \cite{Friston2000} estimates (\autoref{tab:Params}). In fact the results 
are significantly different from virtually every other study. The most obvious discrepancy
is the larger time constants, $\tau_f$, $\tau_s$ and $\tau_0$. 
While of course this could be poor convergence of the algorithm, there is another other possibility.
Unlike all the other methods mentioned, excepting the methods in \autoref{sec:Nonlinear Least Squares},
the algorithm described in \cite{Johnston2008} does not depend on prior distributions.
It is possible then that the bias toward the prior in other methods screwed the results. 
While \cite{Johnston2008} is certainly in the minority; further exhaustive studies 
of the parameters, using unbiased techniques may be called for. A further 
comparison between the distributions found in \cite{Johnston2008} and 
\cite{Friston2000} will be discussed in \autoref{sec:PriorDist}, including
graphs: \autoref{fig:MeanResponseF}, \autoref{fig:MeanResponseJ}.

\section{Conclusion}
There is no ideal solution to solving this system of nonlinear
equations. Exhaustive search type methods such as those employed by
\cite{Johnston2008} and \cite{Vakorin2007} have long
run times even for a single voxel. While Volterra models are an interesting
solution,there have not yet been exhaustive tests to determine whether
such approximations work well throughout state space. The most promising
method of those reviewed here is the Kalman filter based method. It is
able to maintain a fast runtime while still approaching the solution.
While the reliance on a Gaussian estimate to the true posterior distribution could
cause problems, some modifications could make it powerful. 
The particle filter method proposed in the next section bears a strong 
resemblance to the unscented Kalman filter; albeit with more point estimating
the posterior.

### 2010-07-21 doc/thesis/conclusion.tex
removed lots of weasel words...

Although many versions of the BOLD model exist, and it is tempting to use 
more detailed models; from the results found here the issue of bias error
from the BOLD model is not the biggest concern. Clearly clarifying the distributions
of parameters for the prior should be the first concern. 
One future study that would be extremely beneficial in this way  would
be an extensive study of what the priors should truly be. Although \cite{Friston2000}
gives an estimate of what is thought to be reasonable values, and later studies
published their estimate of the distributions, given the interplay between
parameters it is very unlikely that these priors are true to actual distribution
that occurs in vivo. As I mentioned previously, the addition of simultaneous
flow or volume measurements are another potentially powerful way to further confine the model,
and thus deal with the elasticity of parameters. As I mentioned in 
\autoref{sec:BackgroundConclusion}, a chief advantage of using physiologically
plausible models is that such data may in fact be added with relative ease.

In conclusion, using particle filters to estimate the BOLD response are 
an extremely powerful method of fitting to noisy data. The technique also
holds great promise as extensible platform to build more advanced models
and techniques on top of. Integrating further information is necessary to
move beyond the traditional Statistical Parametric Mapping and moving
toward biologically and medically relevant FMRI scanning techniques.

>>>>

Although many versions of the BOLD model exist, and it is tempting to use 
more detailed models; from the results found here the issue of bias error
from the BOLD model is not the biggest concern. Clarifying the distributions
of parameters for the prior should be the first concern; currently no multi-patient
full-volume studies have been done to estimate parameters.
One future study that would be beneficial in this way  would
be an extensive study of what the priors should truly be. Although \cite{Friston2000}
gives an estimate of what is thought to be reasonable values, and later studies
published their estimate of the distributions, given the interplay between
parameters it is unlikely that these priors are true to actual distribution
that occurs in vivo. As I mentioned previously, the addition of simultaneous
flow or volume measurements are another potentially powerful way to further confine the model,
and thus deal with the elasticity of parameters. As I mentioned in 
\autoref{sec:BackgroundConclusion}, a chief advantage of using physiologically
plausible models is that such data may in fact be added with relative ease.

In conclusion, using particle filters to estimate the BOLD response are 
a powerful method of fitting to noisy data. The technique also
holds great promise as extensible platform to build more advanced models
and techniques on top of. Integrating further information is necessary to
move beyond the traditional Statistical Parametric Mapping and moving
toward biologically and medically relevant FMRI scanning techniques.
